---
title: "NHS Borders: Breast Cancer Incidence Report"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
  pdf_document: default
---

```{r, echo = FALSE, warning = FALSE, message = FALSE}

# Load Libraries
library(tidyverse)
library(janitor)
library(infer)
library(plotly)

# Set Colour Scheme

colour_scheme <- c("#092869", "#0391BF", "#6B077B")

# Read In Data

cancer_incidence <- read_csv("data/cancer_incidence_by_health_board.csv") %>% 
  clean_names()

five_year_summary <- read_csv("data/five_year_cancer_incidence_summary.csv") %>% 
  clean_names()

geography_codes <- read_csv("data/geography_codes_and_labels.csv") %>%
  clean_names()
```


```{r, echo = FALSE, warning = FALSE, message = FALSE}

# cancer_incidence Borders Only

cancer_incidence_borders <- cancer_incidence %>% 
  filter(hb == "S08000016") %>% 
  mutate(cancer_site = str_to_title(cancer_site))

# Clean five_year_summary

five_year_summary_borders <- five_year_summary %>% 
  filter(hb == "S08000016")

five_year_summary_borders_incidence <- five_year_summary_borders %>%
  select(id, hb, cancer_site, sex, year | starts_with("incidences"))

five_year_summary_long <- five_year_summary_borders_incidence %>%
  pivot_longer(
    cols = incidences_age_under5:incidences_age85and_over,
    names_to = "age",
    values_to = "incidences"
  ) %>% 
  select(-incidences_all_ages) %>% 
  mutate(age = str_remove(age, "incidences_age"),
         age = str_replace_all(age, "_", ""),
         age = str_replace_all(age, "to", "-"),
         age = case_when(
           age == "under5" ~ "Under 5",
           age == "85andover" ~ "85+",
           TRUE ~ age)) %>% 
  mutate(age = factor(age, levels = c("Under 5", "5-9",
                                      "10-14", "15-19",
                                      "20-24", "25-29",
                                      "30-34", "35-39",
                                      "40-44", "45-49",
                                      "50-54", "55-59",
                                      "60-64", "65-69",
                                      "70-74", "75-79",
                                      "80-84", "85+")))

```


# Introduction

Across all ages and genders, the 3 forms of cancer with the highest number of incidences in NHS Borders between 1997-2021 are:

**Cancer Site / Incidences**

1. Non-melanoma skin cancer /	6174
2. Basal cell carcinoma of the skin	/ 4049
3. Breast	/ 2614

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# cancer_incidence_borders %>% 
#   filter(cancer_site != "All cancer types",
#          sex == "All") %>% 
#   group_by(cancer_site) %>% 
#   summarise(total_incidences = sum(incidences_all_ages)) %>% 
#   arrange(desc(total_incidences)) %>% 
#   head(3)
```


This report will focus on instances of breast cancer, which among females in NHS Borders, has the highest number of incidences, highest mean crude rate and highest mean European age-standardised rate (EASR).

<div class = "row">
  <div class = "col-md-6">
**Cancer Site / Incidences**

**1.** Breast / 2598			
**2.** Non-melanoma skin cancer / 2519			
**3.** Basal cell carcinoma of the skin / 1882	
</div>
</div>

<div class = "row">
<div class = "col-md-6">
**Cancer Site / Mean Crude Rate**

**1.** Breast	/ 179.9348			
**2.** Non-melanoma skin cancer /	174.2811			
**3.** Basal cell carcinoma of the skin /	130.1987	
</div>
</div>

<div class = "row">  
<div class = "col-md-6">
**Cancer Site / Mean EASR**

**1.** Breast / 161.3640			
**2.** Non-melanoma skin cancer / 150.3996			
**3.** Basal cell carcinoma of the skin / 113.9178			
</div>
</div>

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# cancer_incidence_borders %>% 
#   filter(cancer_site != "All cancer types",
#          sex == "Females") %>% 
#   group_by(cancer_site) %>% 
#   summarise(total_incidences = sum(incidences_all_ages)) %>% 
#   arrange(desc(total_incidences)) %>% 
#   head(3)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# cancer_incidence_borders %>% 
#   filter(cancer_site != "All cancer types",
#          sex == "Females") %>% 
#   group_by(cancer_site) %>% 
#   summarise(mean_crude_rate = mean(crude_rate)) %>% 
#   arrange(desc(mean_crude_rate)) %>% 
#   head(3)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# cancer_incidence_borders %>% 
#   filter(cancer_site != "All cancer types",
#          sex == "Females") %>% 
#   group_by(cancer_site) %>% 
#   summarise(mean_easr = mean(easr)) %>% 
#   arrange(desc(mean_easr)) %>% 
#   head(3)
```

# Trends Over Time

*[fig. 1]*

```{r, echo = FALSE, warning = FALSE, message = FALSE}

fig1_plot <- cancer_incidence_borders %>% 
  filter(sex == "Females",
         cancer_site %in% c("All Cancer Types", "Breast")) %>% 
  ggplot() +
  geom_line(aes(x = year, y = incidences_all_ages, colour = cancer_site, group = 1,
                text = paste0("<b>Year:</b> ", year, "<br>", 
                             "<b>Type:</b> ", cancer_site, "<br>",
                             "<b>Incidences:</b> ", incidences_all_ages)),
            size = 1.5) +
  scale_x_continuous(breaks = c(1997:2021)) +
  scale_colour_manual(values = colour_scheme, labels = c("All Combined", "Breast")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  ylim(0, 500) +
  labs(
    x = "\n Year",
    y = "Incidences\n",
    title = "Female Cancer Incidences",
    colour = "Cancer Type:") +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_line(colour = "grey90"))

ggplotly(fig1_plot, tooltip = "text") %>%
  layout(hovermode = "x unified",
         title = list(text = paste0("<b>Female Cancer Incidences</b>",
                                    "<br>",
                                    "<sup>",
                                    "NHS Borders: 1997-2021",
                                    "</sup>")))

```

| Year | No. of Breast Cancer Incidences | % Change From Previous Year   |
|:------|:-------------------|:------------|
| 1997 | 71                | NA         |
| 1998 | 69                | -2.816901  |
| **1999** | **133**               | **92.753623**  |
| 2000 | 69                | -48.120301 |
| 2001 | 81                | 17.391304  |
| **2002** | **131**               | **61.728395**  |
| 2003 | 67                | -48.854962 |
| 2004 | 62                | -7.462687  |
| **2005** | **179**              | **188.709677** |
| 2006 | 68                | -62.011173 |
| 2007 | 55                | -19.117647 |
| **2008** | **154**               | **180.000000** |
| 2009 | 94                | -38.961039 |
| 2010 | 86                | -8.510638  |
| **2011** | **157**              | **82.558140**  |
| 2012 | 103               | -34.394904 |
| 2013 | 114               | 10.679612  |
| **2014** | **130**               | **14.035088**  |
| 2015 | 90                | -30.769231 |
| 2016 | 98                | 8.888889   |
| **2017** | **136**               | **38.775510**  |
| 2018 | 97                | -28.676471 |
| 2019 | 98                | 1.030928   |
| 2020 | 107               | 9.183673   |
| 2021 | 149               | 39.252336  |

**What does this visualisation tell us?**

- That breast cancer incidences in females are driving the trend of all cancer type incidences.
- That there appears to be a historic trend in breast cancer incidences where they peak approximately every 3 years. The average percentage increase in incidences from the previous year at the 7 peaks highlighted is **94%**. 
- That this trend is less evident in 2020 when we may have expected it, indeed there was only a **9%** increase from 2020.
- There is however an increase of **39%** between 2020 and 2021.

# Hypothesis Test

**What does hypothesis testing tell us?**

**Question:** Is the mean number of female breast cancer incidences in *"peak years"* (1999, 2002, 2005, 2008, 2011, 2014, 2017) greater than mean number of female breast cancer incidences in *"non-peak years"* (1997, 1998, 2000, 2001, 2003, 2006, 2007, 2009, 2010, 2012, 2013, 2015, 2016, 2018, 2019)?

**Test Used:** Two Sample Mean Test (Independent)

**Significance Level:** 0.05

**Bootstrap Method:** Permute

**H0:** $\mu{1}$ - $\mu{2}$ = 0
**H1:** $\mu{1}$ - $\mu{2}$ > 0 

Based on a bootstrapped NULL distribution, a very low p-value which is less than 0.05 is returned. We therefor reject H0 in favor of H1 with evidence suggesting that there is a statistically significant increase in the mean number of female breast cancer incidences in *"peak years"*.

In this instance the p-value is returned as 0 which we will interpret as p < 0.001 so as not to be misleading and suggest absolute certainty. As this is lower than our set significance level of 0.05 we can reject H0 which stated that there was no statistically significant difference in the mean number of female breast cancer incidences in *"peak years"* compared to *"non-peak years"*. We reject H0 in favour of H1, whilst not asserting this as true, as their is sufficient evidence to support the Hypothesis that the the mean number of female breast cancer incidences in *"peak years"* is greater than in *"non-peak years"* when considered against the hypothesis that there is no difference.

To explain further the rationale for this, the p-value represents the probability of obtaining the observed statistic (mean difference of female breast cancer incidences 58.04762) from our sample assuming H0 was true (eg. if there was no difference). Therefor the probability of observing this mean difference if there was no difference is very low.

**Why might there be a historic 3 year trend?**

Women who meet screening criteria are invited for breast screening once every 3 years *(NHS National Services Scotland, 2022)*.

**Why might we not see the same peak in 2020 as we may have expected?**

Due to the COVID-19 pandemic, no invites to breast screenings were sent between 30 March 2020 and 3 August 2020 *(Public Health Scotland, 2022)*.

***

# Incidences by Age

*[fig. 2]*

```{r, echo = FALSE, warning = FALSE, message = FALSE}
fig2_plot <- five_year_summary_long %>%
  filter(cancer_site == "Breast",
         sex == "Females") %>%
  ggplot() +
  geom_col(aes(x = age, y = incidences, 
               text = paste0("<b>Age:</b> ", age, "<br>", "<b>Incidences:</b> ", incidences, "<br>")),
           fill = "#0391BF") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  labs(
    x = "\n Age",
    y = "Incidences\n",
    title = "Total Female Breast Cancer Incidences by Age") +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_line(colour = "grey90"))

ggplotly(fig2_plot, tooltip = "text") %>%
  layout(title = list(text = paste0("<b>Total Female Breast Cancer Incidences by Age</b>",
                                    "<br>",
                                    "<sup>",
                                    "NHS Borders: 1997-2021",
                                    "</sup>")))
```

**What does this visualisation tell us?**

- The majority of breast cancer incidences in females appear to be between those aged between 50 and 79.

**Why might these age groups see increased incidence numbers?**

- Currently only women between the ages of 50 and 70 are routinely screened *(NHS National Services Scotland, 2022)*.

<style>
div.yellow {background-color:#fcf8c0; border-radius: 5px; padding: 20px;}

</style>
<div class = "yellow">

**NHS Borders Population Projections:**

Females 50+ 2021: 29889

Females 50+ 2041: 31148 (4.21225% increase)

*(National Records of Scotland, 2023)*

</div>

***

# Conclusions / Recommendations

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

- Screening data should be reviewed to establish if the resulting back-log from COVID-19 has been cleared in order to establish whether a further increase in incidences should be anticipated in 2022.

- Resources should be allocated according to the observed trend of increased incidences every three years

- Research/Analysis should be conducted to further understand and confirm any reason for this trend, including any links to screening schedules.

- Research/Analysis should be conducted to establish whether increased incidence with age is in any way the result of current screening criteria and if therefor screening criteria should be widened.

- Long term service planning should take into consideration the ~4% projected population increase of the female 50-70 demographic in NHS Borders, as rejected by the National Records of Scotland.

</div>

***

# Data Sources

- Public Health Scotland: Incidence by Health Board
https://www.opendata.nhs.scot/dataset/annual-cancer-incidence/resource/3aef16b7-8af6-4ce0-a90b-8a29d6870014

- Public Health Scotland: 5 Year Summary of Incidence by Health Board
https://www.opendata.nhs.scot/dataset/annual-cancer-incidence/resource/e8d33b2b-1fb2-4d59-ad21-20fa2f76d9d5

- Public Health Scotland: Health Board 2014 - Health Board 2019 (Geography Codes)
https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc

# References
- NHS Inform, 2022:
https://www.nhsinform.scot/illnesses-and-conditions/cancer/cancer-types-in-adults/skin-cancer-non-melanoma

- NHS National Services Scotland, 2022: https://www.nss.nhs.scot/specialist-healthcare/screening-programmes/breast-screening/

- National Records of Scotland, 2023:
https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/population/population-projections/sub-national-population-projections/2018-based/detailed-datasets

- Public Health Scotland, 2022:
https://www.publichealthscotland.scot/media/12843/2022-04-26_breast_screening_report.pdf
