---
title: "<b>Breast Cancer Incidence Report</b>"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: "hide"
  pdf_document: default
---

*Disclaimer: This report has been written for the authors learning purposes only.*

```{r, echo = FALSE, warning = FALSE, message = FALSE}

# Load Libraries
library(tidyverse)
library(janitor)
library(infer)
library(plotly)
library(sf)
library(gt)
library(gtExtras)

# Set Colour Scheme

colour_scheme <- c("#092869", "#0391BF", "#6B077B")

# Read In Data

cancer_incidence <- read_csv("data/cancer_incidence_by_health_board.csv") %>% 
  clean_names()

five_year_summary <- read_csv("data/five_year_cancer_incidence_summary.csv") %>% 
  clean_names()

geography_codes <- read_csv("data/geography_codes_and_labels.csv") %>%
  clean_names()

hb_shapefiles <- st_read(dsn = "data/geo_data/", layer = "SG_NHS_HealthBoards_2019", quiet = TRUE)

```

```{r, echo = FALSE, warning = FALSE, message = FALSE}

# Clean Data

## Incidences Figures

cancer_incidence_borders <- cancer_incidence %>% 
  filter(hb == "S08000016") %>% 
  mutate(cancer_site = str_to_title(cancer_site))

## Five Year Summaries

### NHS Borders Only

five_year_summary_borders <- five_year_summary %>% 
  filter(hb == "S08000016")

five_year_summary_borders_incidence <- five_year_summary_borders %>%
  select(id, hb, cancer_site, sex, year | starts_with("incidences"))

five_year_summary_long <- five_year_summary_borders_incidence %>%
  pivot_longer(
    cols = incidences_age_under5:incidences_age85and_over,
    names_to = "age",
    values_to = "incidences"
  ) %>% 
  select(-incidences_all_ages) %>% 
  mutate(age = str_remove(age, "incidences_age"),
         age = str_replace_all(age, "_", ""),
         age = str_replace_all(age, "to", "-"),
         age = case_when(
           age == "under5" ~ "Under 5",
           age == "85andover" ~ "85+",
           TRUE ~ age)) %>% 
  mutate(age = factor(age, levels = c("Under 5", "5-9",
                                      "10-14", "15-19",
                                      "20-24", "25-29",
                                      "30-34", "35-39",
                                      "40-44", "45-49",
                                      "50-54", "55-59",
                                      "60-64", "65-69",
                                      "70-74", "75-79",
                                      "80-84", "85+")))

### All NHS Health Boards

five_year_summary_geo <- five_year_summary %>%
  select(id, hb, cancer_site, sex, year, easr) %>%
  filter(sex == "Females",
         cancer_site == "Breast")

geo_summary <- hb_shapefiles %>%
  left_join(five_year_summary_geo, by = c("HBCode" = "hb"))
  
rm(five_year_summary_borders)
rm(five_year_summary_borders_incidence)
rm(cancer_incidence)

```


# Brief

To inform the planning and provision of cancer treatment services, analyse breast cancer incidence data from the NHS Borders health board. Findings should be presented in a report of 1-2 pages, highlighting any significant insights and identified trends.

# Introduction

Between 1997-2021, incidences of breast cancer in males made up less than 1% of total breast cancer incidences. This report will therefor focus on instances of breast cancer among females.

<div class = "row">
  <div class = "col-md-6">

```{r, message = FALSE}
cancer_incidence_borders %>%
  filter(cancer_site != "All Cancer Types",
         sex == "All") %>%
  group_by(cancer_site) %>%
  summarise(total_incidences = sum(incidences_all_ages)) %>%
  arrange(desc(total_incidences)) %>%
  head(3) %>% 
  gt() %>%
  tab_header(title = "Total Cancer Incidences") %>% 
  cols_label(
    cancer_site = "Cancer Site",
    total_incidences = "Total Incidences") %>%
  tab_options(column_labels.font.weight = "bold",
              table.align = "left")
```

</div>
<div class = "col-md-6">

```{r, message = FALSE}
cancer_incidence_borders %>%
  filter(cancer_site == "Breast",
         sex != "All") %>%
  group_by(sex) %>%
  summarise(total_incidences = sum(incidences_all_ages)) %>%
  arrange(desc(total_incidences)) %>%
  head(3) %>% 
  gt() %>%
  tab_header(title = "Breast Cancer Incidences by Sex") %>% 
  cols_label(
    sex = "Sex",
    total_incidences = "Total Incidences") %>%
  tab_options(column_labels.font.weight = 'bold',
              table.align = "left")
```

</div>
</div>

According to NHS Borders data, breast cancer among females has  the highest number of incidences, highest mean crude rate and highest mean European age-standardised rate (EASR) of any cancer type. See below for a more detail on how the EASR is calculated.

<div class = "row">
<div class = "col-md-6">

```{r, message = FALSE}
cancer_incidence_borders %>%
  filter(cancer_site != "All Cancer Types",
         sex == "Females") %>%
  group_by(cancer_site) %>%
  summarise(total_incidences = sum(incidences_all_ages)) %>%
  arrange(desc(total_incidences)) %>%
  head(3) %>% 
  gt() %>%
  cols_label(
    cancer_site = "Cancer Site",
    total_incidences = "Total Incidences") %>%
  tab_options(column_labels.font.weight = 'bold')
```

</div>
<div class = "col-md-6">

```{r}
cancer_incidence_borders %>%
  filter(cancer_site != "All Cancer Types",
         sex == "Females") %>%
  group_by(cancer_site) %>%
  summarise(mean_easr = mean(easr)) %>%
  arrange(desc(mean_easr)) %>%
  head(3) %>%
  gt() %>%
  cols_label(
    cancer_site = "Cancer Site",
    mean_easr = "Mean EASR") %>%
  tab_options(column_labels.font.weight = 'bold')
```

</div>
</div>

***

# Health Board Comparison

To understand how these rates compare to other health boards in Scotland, we can visualise a five year summary of the EASR - which is a *"weighted mean of the age-specific rates where the weights are taken from the population distribution of a standard population; the ASR is expressed per 100,000." (European Commission, 2023)*

```{r, message = FALSE}

hbc_plot <- geo_summary %>% 
  ggplot(aes(fill = easr, text = paste0("<b>NHS Health Board:</b> ", HBName, "<br>", 
                             "<b>EASR: ~</b>", floor(easr), "<br>"))) + 
  geom_sf(colour = "white", linewidth = 0.04) +
  labs(
    title = "Female Breast Cancer EASR (2017-2021)",
    subtitle = "By NHS Health Board",
    fill = "EASR") +
  scale_fill_distiller(palette = "Blues", direction = +1) +
  theme(panel.background = element_rect(fill = "white"),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())


ggplotly(hbc_plot, tooltip = "text")
```

*NB - Unfortunately data for the individual health boards NHS Western Isles, NHS Shetland and NHS Orkney was not available at the time of report completion.*

```{r, message = FALSE}
five_year_summary %>%
  select(hb, cancer_site, sex, year, easr) %>%
  filter(sex == "Females",
         cancer_site == "Breast",
         hb != "GR0800001") %>%
  left_join(geography_codes, "hb") %>% 
  select(hb_name, year, sex, cancer_site, easr) %>% 
  arrange(desc(easr)) %>% 
  gt() %>%
  cols_label(
    hb_name = "NHS Health Board",
    year = "Year(s)",
    sex = "Sex",
    cancer_site = "Cancer Site",
    easr = "EASR") %>% 
  tab_options(column_labels.font.weight = 'bold') %>%
  data_color(columns = easr, palette = "Blues")
  
```

***

# Trends Over Time

*[fig. 1]*

```{r, message = FALSE, warning = FALSE, fig.width = 10}

fig1_plot <- cancer_incidence_borders %>% 
  filter(sex == "Females",
         cancer_site %in% c("All Cancer Types", "Breast")) %>% 
  ggplot() +
  geom_line(aes(x = year, y = incidences_all_ages, colour = cancer_site, group = 1,
                text = paste0("<b>Year:</b> ", year, "<br>", 
                             "<b>Type:</b> ", cancer_site, "<br>",
                             "<b>Incidences:</b> ", incidences_all_ages)),
            size = 1.5) +
  scale_x_continuous(breaks = c(1997:2021)) +
  scale_colour_manual(values = colour_scheme, labels = c("All Combined", "Breast")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  ylim(0, 500) +
  labs(
    x = "\n Year",
    y = "Incidences\n",
    title = "Female Cancer Incidences",
    colour = "Cancer Type:") +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_line(colour = "grey90"))

ggplotly(fig1_plot, tooltip = "text") %>%
  layout(hovermode = "x unified",
         title = list(text = paste0("<b>Female Cancer Incidences</b>",
                                    "<br>",
                                    "<sup>",
                                    "NHS Borders: 1997-2021",
                                    "</sup>")))

```

**What does this visualisation tell us?**

- That breast cancer incidences in females appear to be driving the trend of all cancer type incidences.
- That breast cancer incidences appear to peak approximately every 3 years.

When we look at the year-on-year percentage changes in breast cancer incidences we can gain further insights. The table below shows:

- The average percentage increase in incidences from the previous year at the 8 peaks highlighted is **84%**. 
- This trend is less evident in 2020 when we may have expected it, indeed there was only a **9%** increase from 2020.

**Why might there be a 3 year trend?**

Women who meet screening criteria are invited for breast screening once every 3 years *(NHS National Services Scotland, 2022)*.

**Why might we not see the same peak in 2020 as we may have expected?**

Due to the COVID-19 pandemic, no invites to breast screenings were sent between 30 March 2020 and 3 August 2020 *(Public Health Scotland, 2022)*.

```{r, message = FALSE}

cancer_incidence_borders %>% 
  filter(sex == "Females",
         cancer_site == "Breast") %>% 
  select(year, sex, cancer_site, incidences_all_ages) %>% 
  mutate(yearly_pct_change = round((incidences_all_ages - lag(incidences_all_ages)) / lag(incidences_all_ages) * 100)) %>% 
  gt() %>%
  cols_label(
    year = "Year",
    sex = "Sex",
    cancer_site = "Cancer Site",
    incidences_all_ages = "No. of Incidences",
    yearly_pct_change = "% Change from Previous Year") %>% 
  tab_options(column_labels.font.weight = 'bold') %>% 
  gt_highlight_rows(rows = c(3, 6, 9, 12, 15, 18, 21, 24),
                    bold_target_only = TRUE,
                    target_col = yearly_pct_change)
  
```

***

# Hypothesis Test

**Question:** Is the greater number of female breast cancer incidences in *"peak years"* (1999, 2002, 2005, 2008, 2011, 2014, 2017) compared to *"non-peak years"* (1997, 1998, 2000, 2001, 2003, 2006, 2007, 2009, 2010, 2012, 2013, 2015, 2016, 2018, 2019) statistically significant?

```{r, message = FALSE, warning = FALSE}
cancer_incidence_borders_sample <- cancer_incidence_borders %>%
  filter(sex == "Females", cancer_site == "Breast") %>%
  select(id, cancer_site, sex, year, incidences_all_ages) %>% 
    mutate(peak = case_when(
    year == 1999 ~ "peak",
    year == 2002 ~ "peak",
    year == 2005 ~ "peak",
    year == 2008 ~ "peak",
    year == 2011 ~ "peak",
    year == 2014 ~ "peak",
    year == 2017 ~ "peak",
    TRUE          ~ "standard"
    )
  )

observed_stat <- cancer_incidence_borders_sample %>% 
  specify(incidences_all_ages ~ peak) %>%
  calculate(stat = "diff in means", order = c("peak", "standard"))

null_distribution <- cancer_incidence_borders_sample %>% 
  specify(response = incidences_all_ages, explanatory = peak) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("peak", "standard"))

p_value <- null_distribution %>%
  get_p_value(obs_stat = observed_stat, direction = "right")

```

<div class = "row">
  <div class = "col-md-6">

**Test Used:** Two Sample Mean Test (Independent)<br>
**Significance Level:** 0.05

  </div>
<div class = "col-md-6">

**H0:** $\mu{1}$ - $\mu{2}$ = 0<br>
**H1:** $\mu{1}$ - $\mu{2}$ > 0 
		
</div>
</div>

**Result:** Based on a bootstrapped NULL distribution, a very low p-value which is less than 0.05 is returned. We therefor reject H0 in favor of H1 with evidence suggesting that there is a statistically significant increase in the mean number of female breast cancer incidences in *"peak years"*.

***

# Incidences by Age

*[fig. 2]*

```{r, message = FALSE, warning = FALSE, fig.width = 8}
fig2_plot <- five_year_summary_long %>%
  filter(cancer_site == "Breast",
         sex == "Females") %>%
  ggplot() +
  geom_col(aes(x = age, y = incidences, 
               text = paste0("<b>Age:</b> ", age, "<br>", "<b>Incidences:</b> ", incidences, "<br>")),
           fill = "#0391BF") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  labs(
    x = "\n Age",
    y = "Incidences\n",
    title = "Total Female Breast Cancer Incidences by Age") +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_line(colour = "grey90"))

ggplotly(fig2_plot, tooltip = "text") %>%
  layout(title = list(text = paste0("<b>Total Female Breast Cancer Incidences by Age</b>",
                                    "<br>",
                                    "<sup>",
                                    "NHS Borders: 1997-2021",
                                    "</sup>")))
```

**What does this visualisation tell us?**

- The majority of breast cancer incidences in females appear to be between those aged between 50 and 79.

**Why might these age groups see increased incidence numbers?**

- Currently only women between the ages of 50 and 70 are routinely screened *(NHS National Services Scotland, 2022)*.

<style>
div.yellow {background-color:#fcf8c0; border-radius: 5px; padding: 20px;}

</style>
<div class = "yellow">

**NHS Borders Population Projections:**

Females 50+ 2021: 29889

Females 50+ 2041: 31148 (4.21225% increase)

*(National Records of Scotland, 2023)*

</div>

***

# Conclusions / Recommendations

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

- Screening data should be reviewed to establish if the resulting back-log from COVID-19 has been cleared in order to establish whether a further increase in incidences should be anticipated in 2022.

- Resources should be allocated according to the observed trend of increased incidences every three years

- Research/Analysis should be conducted to further understand and confirm any reason for this trend, including any links to screening schedules.

- Research/Analysis should be conducted to establish whether increased incidence with age is in any way the result of current screening criteria and if therefor screening criteria should be widened.

- Long term service planning should take into consideration the ~4% projected population increase of the female 50-70 demographic in NHS Borders, as projected by the National Records of Scotland.

</div>

***

# Data Sources

- SpatialData.gov.scot Metadata Portal: NHS Scotland Health Boards
https://spatialdata.gov.scot/geonetwork/srv/api/records/f12c3826-4b4b-40e6-bf4f-77b9ed01dc14

- Public Health Scotland: 5 Year Summary of Incidence by Health Board
https://www.opendata.nhs.scot/dataset/annual-cancer-incidence/resource/e8d33b2b-1fb2-4d59-ad21-20fa2f76d9d5

- Public Health Scotland: Geography Codes and Labels
https://www.opendata.nhs.scot/dataset/geography-codes-and-labels

- Public Health Scotland: Incidence by Health Board
https://www.opendata.nhs.scot/dataset/annual-cancer-incidence/resource/3aef16b7-8af6-4ce0-a90b-8a29d6870014

# References

- European Commission, 2023
https://ecis.jrc.ec.europa.eu/info/glossary.html

- NHS National Services Scotland, 2022: https://www.nss.nhs.scot/specialist-healthcare/screening-programmes/breast-screening/

- National Records of Scotland, 2023:
https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/population/population-projections/sub-national-population-projections/2018-based/detailed-datasets

- Public Health Scotland, 2022:
https://www.publichealthscotland.scot/media/12843/2022-04-26_breast_screening_report.pdf
